Basic Model formulation

### 2015-02-02 documentation/model_formulation.Rmd
develop pair-approximation ODE system, solve ODE

This is assuming that the local environmental suitability is enhanced by a local facilitation term similar to the implementation in Kéfi et al 2007. Thus, the reduction in growth rate, $r$, by aridity, $a$, is compensated by a function of $q_{0|1}$, the local density of cells in state 1 (i.e. vegetation) given that the focal cell is in state 0 (i.e. empty), and maximizes to one if the cell has four neighbors:

$$ a = a_0 + (f - a_0) q_{0|1} \,,$$

@@

$$ K = K_0 (1 - c q_{0|1})$$

@@

Mortality becomes enhanced in sparse environments by patches that are attracting grazers. We assume this attractivity enhances search efficiency locally by the value $v$

$$ b = b_0  + v q_{1|1} $$

@@

$$ h = h_0  / (1 - p  q_{1|1}) $$

@@

@@

@@

The vegetation dynamics of the system can be fully described by two differential equations (omitting indexed zeros of equations for easier reading):

$$ \frac{\mathrm{d}\rho_1}{\mathrm{d}t} = r ( a + (f - a) q_{0|1} ) \rho_1^{1 + \alpha} (1-\frac{\rho_1}{K (1 - c q_{0|1})}) - m \rho_1 - \frac{(b  + v q_{1|1} )\rho_1^{1+q}L}{1+( b  + v q_{1|1} ) ( h  / (p  q_{1|1}))\rho_1^{1+q}} $$

$$ \frac{\mathrm{d}\rho_0}{\mathrm{d}t} =  m \rho_1 + \frac{(b  + v q_{1|1} )\rho_1^{1+q}L}{1+( b  + v q_{1|1} ) ( h  / (p  q_{1|1}))\rho_1^{1+q}} - r ( a + (f - a) q_{0|1} ) \rho_1^{1 + \alpha} (1-\frac{\rho_1}{K (1 - c q_{0|1})})$$

Next steps:

- formulate equations for $\frac{\mathrm{d}\rho_{0,1}}{\mathrm{d}t}$, $\frac{\mathrm{d}\rho_{0,0}}{\mathrm{d}t}$, and $\frac{\mathrm{d}\rho_{1,1}}{\mathrm{d}t}$

>>>>

This is assuming that the local environmental suitability is enhanced by a local facilitation term similar to the implementation in Kéfi et al 2007. Thus, the reduction in growth rate, $r$, by aridity, i.e. the inverse of environmental quality $b$, is compensated by a function of $q_{0|1}$, the local density of cells in state 1 (i.e. vegetation) given that the focal cell is in state 0 (i.e. empty), and maximizes to one if the cell has four neighbors:

$$ b = b^* + (f - b^*) q_{0|1} \,,$$

@@

$$ K = K^* (1 - c q_{0|1})$$

@@

Mortality becomes enhanced in sparse environments by patches that are attracting grazers. We assume this attractivity enhances search efficiency, $a$, locally by the value $v$

$$ a = a^*  + v q_{1|1} $$

@@

$$ h = h^*  / (1 - p  q_{1|1}) $$

@@

the transition probabilities therefore become

$$ w_{1,0} = m + ( a^*  + v q_{1|1})L\rho_1^{q}{1+( a^*  + v q_{1|1})(h^*/ (1 - p  q_{1|1}))\rho_1^{1+q}} $$

$$ w_{0,1} = \frac{r \rho_1^{1+\alpha} (b^* + (f - b^*) q_{0|1}) (1-\frac{\rho_1}{K^* (1 - c q_{0|1})})}{1-\rho_1} $$

@@

We can use the transition probabilities of cells  to describe the full system of differential equations.

$$  \frac{\mathrm{d}\rho_{11}}{\mathrm{d}t} = 2 \rho_{10}w_{0,1} - 2 \rho_{11} w_{1,0}  $$

$$  \frac{\mathrm{d}\rho_{1}}{\mathrm{d}t} = \rho_{0}w_{0,1} - \rho_{1}w_{1,0}   $$

$$  \frac{\mathrm{d}\rho_{10}}{\mathrm{d}t} = 2 \rho_{00}w_{0,1} + 2 \rho_{11}w_{1,0} - \rho_{10}w_{0,1}    $$

$$  \frac{\mathrm{d}\rho_{0}}{\mathrm{d}t} = \rho_{1}w_{1,0} - \rho_{0}w_{0,1} $$

$$  \frac{\mathrm{d}\rho_{00}}{\mathrm{d}t} = 2 \rho_{10}w_{1,0} - 2 \rho_{00}w_{0,1}  $$

Due to the conservation equations, two of these are sufficient to fully describe the dynamics of the system.

We chose to go with the first two and substitute

$$  \frac{\mathrm{d}\rho_11}{\mathrm{d}t} = 2 \rho_{10} \frac{r \rho_1^{1+\alpha} (b^* + (f - b^*) q_{0|1}) (1-\frac{\rho_1}{K^* (1 - c q_{0|1})})}{1-\rho_1} - 2 \rho_{11} m -  2 \rho_{11} \frac{(a^*  + v q_{1|1})L\rho_1^{q}}{1+(a^*  + v q_{1|1})(h^*  / (1 - p  q_{1|1}))\rho_1^{1+q}}  $$

$$  \frac{\mathrm{d}\rho_{1}}{\mathrm{d}t} = \rho_{0}\frac{r \rho_1^{1+\alpha} (b^* + (f - b^*) q_{0|1}) (1-\frac{\rho_1}{K^* (1 - c q_{0|1})})}{1-\rho_1} - \rho_{1}m - \frac{(a^*  + v q_{1|1})L\rho_1^{1+q}}{1+(a^*  + v q_{1|1})(h^*  / (1 - p  q_{1|1}))\rho_1^{1+q}}   $$

they can be simplified to

$$  \frac{\mathrm{d}\rho_11}{\mathrm{d}t} = 2 \rho_{10} \frac{r \rho_1^{1+\alpha} (b^* + (f - b^*) q_{0|1}) (1-\frac{\rho_1}{K^* (1 - c q_{0|1})})}{1-\rho_1} - 2 \rho_{11} m -  2 \rho_{11} \frac{(a^*  + v q_{1|1})L\rho_1^{q}}{1+(a^*  + v q_{1|1})(h^*  / (1 - p  q_{1|1}))\rho_1^{1+q}}  $$

$$  \frac{\mathrm{d}\rho_{1}}{\mathrm{d}t} = r \rho_1^{1+\alpha} (b^* + (f - b^*) q_{0|1}) (1-\frac{\rho_1}{K^* (1 - c q_{0|1})}) - \rho_{1}m - \frac{(a^*  + v q_{1|1})L\rho_1^{1+q}}{1+(a^*  + v q_{1|1})(h^*  / (1 - p  q_{1|1}))\rho_1^{1+q}}   $$

as a matter of fact the equation for $\frac{\mathrm{d}\rho_{1}}{\mathrm{d}t}$ will re-convert into the regular population equation.

@@

thus,

$$ q_{0|1} = \frac{\rho_{10}}{\rho_0} =  \frac{\rho_{1} - \rho_{11}}{1 - \rho_1} $$

$$ q_{1|1} = \frac{\rho_{11}}{\rho_1} $$

and substituting yields

$$  \frac{\mathrm{d}\rho_{11} }{\mathrm{d}t} = 2 (\rho_1 - \rho_{11}) \frac{r \rho_1^{1+\alpha} (b^* + (f - b^*) \frac{\rho_{1} - \rho_{11}}{1 - \rho_1}) (1-\frac{\rho_1}{K^* (1 - c \frac{\rho_{1} - \rho_{11}}{1 - \rho_1})})}{1-\rho_1} - 2 \rho_{11} m -  2 \rho_{11} \frac{(a^*  + v \frac{\rho_{11}}{\rho_1})L\rho_1^{q}}{1+(a^*  + v \frac{\rho_{11}}{\rho_1})(h^*  / (1 - p  \frac{\rho_{11}}{\rho_1}))\rho_1^{1+q}}  $$

$$  \frac{\mathrm{d}\rho_{1}}{\mathrm{d}t} = r \rho_1^{1+\alpha} (b^* + (f - b^*) \frac{\rho_{1} - \rho_{11}}{1 - \rho_1}) (1-\frac{\rho_1}{K^* (1 - c \frac{\rho_{1} - \rho_{11}}{1 - \rho_1})}) - \rho_{1}m -  \frac{(a^*  + v \frac{\rho_{11}}{\rho_1})L\rho_1^{1+q}}{1+(a^*  + v \frac{\rho_{11}}{\rho_1})(h^*  / (1 - p  \frac{\rho_{11}}{\rho_1}))\rho_1^{1+q}}   $$

Thus, we can solve the ODE for a given set of parameters

```{r}

parms <- list(
    m = 0.05,  # intrinsic mortality of plants 
    r = 1,   # growth rate
    f = 0.7, #  local facilitation
    b = 0.3,  # environmental quality inverse of aridity
    K = 0.9,  # carrying capacity
    alpha = 0.1, # water runoff
    c = 0.2, # local competition
    a = 0.6, # search efficiency
    v = 0.1, # attraction effect
    h = 50, # handling time (one individual on landscape unit)
    p = 0.2, # protection effect
    L = 4, # livestock units per landscape unit 
    q = 1 # search ineffiency at low cover
  )

we get the following dynamics

rho <- c(rho_1 = 0.9, rho_11 = 0.8, rho_10 = 0.1, rho_00 = 0, rho_0 = 0.1)

d_rho_1 <- function(rho, parms) { 
  with(parms,
       r * (b + (f - b) * (rho[1] - rho[2])/(1- rho[1]) ) * rho[1]^( 1 + alpha) * (1 - rho[1]/(K * (1-c*(rho[1] - rho[2])/(1- rho[1])) ) ) - m * rho[1] - ( (a + v*rho[2]/rho[1]) * rho[1]^( 1 + q) * L)/(1 +(a + v*rho[2]/rho[1]) * (h / (1- p * rho[2]/rho[1]) ) * rho[1]^( 1 + q)) 
       )  
  }

d_rho_11 <- function(rho,  parms) { 
  with(parms,
       2* (rho[1] - rho[2]) * r * (b + (f - b) * (rho[1] - rho[2])/(1- rho[1]) ) * rho[1]^( 1 + alpha) * (1 - rho[1]/(K * (1-c*(rho[1] - rho[2])/(1- rho[1])) ) ) / (1-rho[1]) - 2 * rho[2] * m  - 2 * rho[2] * ( (a + v*rho[2]/rho[1]) * rho[1]^( 1 + q) * L)/(1 +(a + v*rho[2]/rho[1]) * (h / (1- p * rho[2]/rho[1]) ) * rho[1]^( 1 + q)) 
       )  
  }

odesys_spex <- function(t, rho, parms = model_parms) {
  rho_1 <- d_rho_1(rho, parms)
  rho_11 <- d_rho_11(rho, parms)
  list(c(rho_1, rho_11, rho_1-rho_11, 1-rho_1-(rho_1-rho_11),1-rho_1  
               ) )
}

# running the ode-solver
ODE_spex <- ode(y = rho, func = odesys_spex, times = 1.1^(1:50), parms = parms)

# transfer into ouput and calculate missing rho values
ODE_spex <- as.data.frame(ODE_spex )

plot(rho_1 ~ time , data = ODE_spex , type = "l", ylim = c(0,1))
axis(4, at = round(tail(ODE_spex,1)$rho_1,2),cex.axis = 0.7, las = 1 )

Now, we can also plot the clustering coefficient of this landscape, being

$$ c_11 = q_{1|1}/\rho_1 = \frac{\rho_11}{\rho_1^2}$$

```{r, echo=FALSE}
plot(I(rho_11/rho_1^2) ~ time , data = ODE_spex , type = "l", ylim = c(0,2))
